Temporal dataflow graphs and their application to generative music

### 2017-03-22 article-icmc2017.tex
save

    An example would be an audio filter : $p_1$ is a low-pass and $p_2$ a reverb.

    The simplest strategy can be to only allow execution when all processes are active and have been given a definite order of execution : in this case the processes will only execute from $t=3$ to $t=5$ and behave like function composition. 
    While safe from the point of view of software execution, we will show that other execution policies leveraging information from the temporal layout of the processes can provide composers with new creative capabilities.

    The main problem that arises is that composers may specify relationships between processes, but such processes may not always be active. 
    We start by making the case for an environment associated with the dataflow graph that contains the values of the inputs and outputs.
    Then, since a graph may not always be fully active, we show that it is also meaningful to have implicit connections between nodes of the graph, that will use the environment.
    This allows dynamic routing of the processes according to their temporal order during execution, and extends routing to pattern-matched elements of the environment instead of single values.
    We enumerate four different behaviours regarding the input and output parameters when the nodes are implicitely connected at runtime using the environment : strict, glutton, parallelized, and replaced. 
    Additionnaly, when the node are explicitely connected, we introduce a delay mechanism that allows a computation to happen on data previously produced by another node of the graph.
    \subsection{Related works}
    Citer Kyma
    Céu
    Perl

@@

	\section{Graph definitions}
    We rely on two families of graphs for the execution of the software : 
    \begin{itemize}
        \item A dataflow graph between processes : processes have input and output ports that can be connected together to enforce data dependencies.
        \item A temporal graph, based on the OSSIA formalism: $G_t(V_t, E_t)$ where $V_t$ is the set of time nodes and $E_t$ is the set of time constraints in a score.
    \end{itemize}
    There may be more than a single temporal graph : they can be arranged hierarchically.

@@

    At each tick, the state of each temporal graph is updated with the following informations : 
    \begin{itemize}
        \item Which processes are currently running.
        \item For how long these processes have been running.
    \end{itemize}

@@

    Event-driven vs time-driven vs sound-driven ; each node has a "driver" port which is the time or the sound. (Do like LV2 : pass empty buffers). Pb : multi-rate ... voir méthode pour passer de multi-rate à single rate dans PLOP2006 et Signal Flow Graphs and Data Flow Graphs (3.2, 3.3), dans Handbook of Signal Processing (First Edition)

    % Pour thèse : essayer de spécifier l'ensemble des cas possible / configuraitons d'activation de manière statique à partir d'un scénario donné, même si c'est exponentiel..

@@

    We extend this with a permanent environment, that contains a mapping of key-value pairs. 
    Keys are adresses : \lstinline|/foo/bar| and values can be any kind of data : numbers, strings, etc. 
    The values of the environment may vary independently of the execution of the program : for instance elements of the environment could be changed through the network, graphical or physical controls.

    We then have to provide a mean to use the values of this environment to the dataflow graph.
    Thus, we extend the inputs and outputs ports with the following expliciteness semantic :

@@

    During a tick, adresses have access to two sets of values : 
    \begin{itemize}
        \item The values that were in the environment at the beginning of the tick.
        \item The values produced by previous nodes in the graph.
    \end{itemize}

@@

    This mechanism allows for new patterns during the creation of a dataflow program : for instance, instead of accepting data from a single source, a port could specify a pattern (see OSC 1.1 paper with XPath) such as : \lstinline|/foo.*/bar/b[a-zA-Z]|. 
    % TODO pattern-matched addresses on the output : seems hard ? how to match it with a "pattern-matched" input port ??
	%Inner data : reduces the possibilities of connection : everything has to be there (explicit)

	%Outer data : have to care about overwriting (implicit)
    %Implicit edges allow for features such as pattern-matching : one node outputs to random, unknown other inputs, or to the global state.
    % For inputs to handle this, we have to provide them with a "can accept" function.

@@

    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

@@

    We will get the following set of equations representing the behaviour : 
    \begin{itemize}
        \item During $t_0$ : $\emptyset$.
        \item During $t_1$ : $f_2 \circ f_1 $ or $f_1 \circ f_2$.
        \item During $t_2$ : $\emptyset$. 
    \end{itemize}

@@

    If we take the same case than previously, the behaviour is :
    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_2 \circ f_1$ or $f_1 \circ f_2$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

@@

	The bufferisation can behave in two way :
	\begin{itemize}
		\item Readers of the buffer always start from the same point (the beginning of the previous function in the callback chain).
		\item Readers of the buffer continue from the latest read position.
		Question : what happens in case of multiple read ?
	\end{itemize}

    Note : this question has an answer in PLOP2006.

@@

    We propose for instance : 
    \begin{itemize}
        \item For simple parameters such as MIDI control changes or OSC messages : the latest value replaces the previous ones.
        \item For audio streams : the streams are mixed together.
        \item For MIDI notes : the notes are played together.
    \end{itemize}

@@

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_1$ and $f_2$ run in parallel, with the same value for $/a$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

@@

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_1$ or $f_2$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

@@

    Hierarchical : cable, port, node, address, global.

    Default global behaviour : glutton.

@@

    That is, if the execution happens in the order of the fig.~\ref{fig.simple}, we get the following execution :

    \begin{itemize}
        \item $t_0$ : $f_1$.
        \item $t_1$ : $f_2 \circ f_1$.
        \item $t_2$ : $f_2$. 
    \end{itemize}

@@

	Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration : see PLOP 2006 article (multi-stream, etc), sample accuracy, etc

    Explicit cables : optimization.
\end{document}

>>>>

    An example would be an audio filter: $p_1$ is a low-pass and $p_2$ a reverb.

    The simplest strategy can be to only allow execution when all processes are active and have been given a definite order of execution: in this case the processes will only execute from $t=3$ to $t=5$ and behave like function composition. 
    While safe from the point of view of software execution, we will show that other execution policies leveraging information from the temporal layout of the processes can provide composers with new creative capabilities.

    The main problem that arises is that composers may specify relationships between processes, but such processes may not always be active. 
    We start by making the case for an environment associated with the dataflow graph that contains the values of the inputs and outputs.
    Then, since a graph may not always be fully active, we show that it is also meaningful to have implicit connections between nodes of the graph, that will use the environment.
    This allows dynamic routing of the processes according to their temporal order during execution, and extends routing to pattern-matched elements of the environment instead of single values.
    We enumerate four different behaviours regarding the input and output parameters when the nodes are implicitely connected at runtime using the environment: strict, glutton, parallelized, and replaced. 
    Additionnaly, when the node are explicitely connected, we introduce a delay mechanism that allows a computation to happen on data previously produced by another node of the graph.
    \subsection{Related works}
    Citer Kyma
    Céu
    Perl

@@

	\section{Graph definitions}
    We rely on two families of graphs for the execution of the software: 
    \begin{itemize}
        \item A dataflow graph between processes: processes have input and output ports that can be connected together to enforce data dependencies.
        \item A temporal graph, based on the OSSIA formalism: $G_t(V_t, E_t)$ where $V_t$ is the set of time nodes and $E_t$ is the set of time constraints in a score.
    \end{itemize}
    There may be more than a single temporal graph: they can be arranged hierarchically.

@@

    At each tick, the state of each temporal graph is updated with the following informations: 
    \begin{itemize}
        \item Which processes are currently running.
        \item For how long these processes have been running.
    \end{itemize}

@@

    Event-driven vs time-driven vs sound-driven ; each node has a "driver" port which is the time or the sound. (Do like LV2: pass empty buffers). Pb: multi-rate ... voir méthode pour passer de multi-rate à single rate dans PLOP2006 et Signal Flow Graphs and Data Flow Graphs (3.2, 3.3), dans Handbook of Signal Processing (First Edition)

    % Pour thèse: essayer de spécifier l'ensemble des cas possible / configuraitons d'activation de manière statique à partir d'un scénario donné, même si c'est exponentiel..

@@

    We extend this with a permanent environment, that contains a mapping of key-value pairs. 
    Keys are adresses: \lstinline|/foo/bar| and values can be any kind of data: numbers, strings, etc. 
    The values of the environment may vary independently of the execution of the program: for instance elements of the environment could be changed through the network, graphical or physical controls.

    We then have to provide a mean to use the values of this environment to the dataflow graph.
    Thus, we extend the inputs and outputs ports with the following expliciteness semantic:

@@

    During a tick, adresses have access to two sets of values: 
    \begin{itemize}
        \item The values that were in the environment at the beginning of the tick.
        \item The values produced by previous nodes in the graph.
    \end{itemize}

@@

    This mechanism allows for new patterns during the creation of a dataflow program: for instance, instead of accepting data from a single source, a port could specify a pattern (see OSC 1.1 paper with XPath) such as: \lstinline|/foo.*/bar/b[a-zA-Z]|. 
    % TODO pattern-matched addresses on the output: seems hard ? how to match it with a "pattern-matched" input port ??
	%Inner data: reduces the possibilities of connection: everything has to be there (explicit)

	%Outer data: have to care about overwriting (implicit)
    %Implicit edges allow for features such as pattern-matching: one node outputs to random, unknown other inputs, or to the global state.
    % For inputs to handle this, we have to provide them with a "can accept" function.

@@

     NOTE : if we have cables  we shouldn't have the "parameters" spill to other nodes. 
    \subsection{Strict}
    An execution of a sub-graph can only happen when all the nodes of the sub-graph are executing.

@@

    We will get the following set of equations representing the behaviour: 
    \begin{itemize}
        \item During $t_0$: $\emptyset$.
        \item During $t_1$: $f_2 \circ f_1 $ or $f_1 \circ f_2$.
        \item During $t_2$: $\emptyset$. 
    \end{itemize}

@@

    If we take the same case than previously, the behaviour is:
    \begin{itemize}
        \item $t_0$: $f_1$.
        \item $t_1$: $f_2 \circ f_1$ or $f_1 \circ f_2$.
        \item $t_2$: $f_2$. 
    \end{itemize}

@@

	The bufferisation can behave in two way:
	\begin{itemize}
		\item Readers of the buffer always start from the same point (the beginning of the previous function in the callback chain).
		\item Readers of the buffer continue from the latest read position.
		Question: what happens in case of multiple read ?
	\end{itemize}

    Note: this question has an answer in PLOP2006.

@@

    We propose for instance: 
    \begin{itemize}
        \item For simple parameters such as MIDI control changes or OSC messages: the latest value replaces the previous ones.
        \item For audio streams: the streams are mixed together.
        \item For MIDI notes: the notes are played together.
    \end{itemize}

@@

    \begin{itemize}
        \item $t_0$: $f_1$.
        \item $t_1$: $f_1$ and $f_2$ run in parallel, with the same value for $/a$.
        \item $t_2$: $f_2$. 
    \end{itemize}

@@

    \begin{itemize}
        \item $t_0$: $f_1$.
        \item $t_1$: $f_1$ or $f_2$.
        \item $t_2$: $f_2$. 
    \end{itemize}

@@

    Hierarchical: cable, port, node, address, global.

    Default global behaviour: glutton.

@@

    That is, if the execution happens in the order of the fig.~\ref{fig.simple}, we get the following execution:

    \begin{itemize}
        \item $t_0$: $f_1$.
        \item $t_1$: $f_2 \circ f_1$.
        \item $t_2$: $f_2$. 
    \end{itemize}

@@

	Divide the time in abstract segment according to the activation graph; gives us a set of equations for each variable.
	\section{Implementation}
    \section{Conclusion}
    Additional consideration: see PLOP 2006 article (multi-stream, etc), sample accuracy, etc

    Explicit cables: optimization.
\end{document}
